Dialect: A Compact Line Notation Based on SMILES

### 2022-03-24 manuscript.tex
Standardizes placement of references. closes #99

Simplified Molecular Input Line Entry System (SMILES) was first described by Weininger in 1988.\cite{weininger:1988} As a line notation,\cite{wiswesser:1968} SMILES represents molecules as single line character sequences, or strings. SMILES strings can be read and written algorithmically, and can be canonicalized\cite{weininger:1989,oBoyle:2012} to yield unique molecular identifiers.

SMILES has since been widely adopted. Read and write functionalities are routinely supported by popular cheminformatics toolkits, including: Open Babel;\cite{oBoyle:2011} RDKit;\cite{rdkit} Chemistry Development Kit;\cite{steinbeck:2003} JChem Base;\cite{jchemBase} the Daylight Toolkit;\cite{daylightToolkit} and OEChem TK.\cite{oetk} SMILES encodings can be found in many public-facing databases, including: PubChem;\cite{kim:2016} ChEBI;\cite{degtyarenko:2007} ZINC;\cite{irwin:2005} ChEMBL;\cite{gaulton:2012} and Wikipedia.\cite{wikipediaCAS} Raw SMILES strings have been used extensively in machine learning applications.\cite{sousa:2021} SMILES has also been extended to carry various forms of metadata, as exemplified by Jmol SMILES,\cite{hanson:2016} CurleySMILES,\cite{drefahl:2011} BigSMILES,\cite{lin:2019} and CXSMILES.\cite{cxsmiles}

@@

Earlier descriptions of SMILES were published in various media, although these are not cited by Weininger. An currently-accessible example is a 1987 report to the US Environmental Protection Agency.\cite{anderson:1987} Although the language described there shares many features with later descriptions, there are important differences as well.

Since 1988, two authoritative sources have attempted to clarify or revise Weininger's 1988 description of SMILES. A 2003 book chapter by Weininger addressed some of the limitations of the paper.\cite{weininger:2008} The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc. ("Daylight"), recapitulates previously-published material.\cite{daylightTheory}. The lack of other written sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication.\cite{gasteiger:2018}

In 2007 a documentation effort that would become known as OpenSMILES began.\cite{openSMILES} OpenSMILES was conceived as "a non-proprietary specification for the SMILES language." Many of the points left open by previous SMILES documentation efforts were addressed. Noteworthy contributions included: the first formal grammar, which was later adapted;\cite{denjoed:2010} many refinements around stereochemistry; and the introduction of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count. OpenSMILES also left several points of syntactic and semantic ambiguity unsettled, while introducing extensions of its own.

In 2019 IUPAC announced the SMILES+ initiative.\cite{smilesPlusSpecification} Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to "establish a formalized recommended up-to-date specification of the SMILES format." SMILES+ took as its starting point the documentation produced by the OpenSMILES project. Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted.\cite{smilesPlusGitHub}

@@

The lack of a detailed, comprehensive, and universally-applied SMILES specification also hampers broader data integrity efforts. The FAIR Guiding Principles\cite{wilkinson:2016} identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability. Interoperability in particular requires that "(meta)data" use a formal, accessible, shared, and broadly applicable language for knowledge representation." Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.

This situation has persisted long enough that the notion of a SMILES "standard" may not currently be feasible. A dozen or more slightly different implementations are now in use. None of them can be considered a reference standard. Some have introduced extensions not recognized by the others, contributing to language "drift" while compounding the already formidable difficulties faced by implementors. Others claim SMILES compatibility while ignoring certain aspects of written documentation. The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use.\cite{daylightToolkit} For several years Daylight operated a publicly-accessible web service based on its toolkit that depicted SMILES strings, but it has since been decommissioned.\cite{depictArchive} Although SMILES may underpin modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.

@@

Dialect's system of representation is based on the \textit{molecular graph} concept.\cite{balaban:1985} A graph is a data structure comprised of a set of nodes (which map to "atoms") and a set of pairwise relationships between them called edges (which map to "bonds"). A molecular graph is a specialized graph onto which atomic and bonding metadata have been overlaid. Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language\cite{murray-rust:2011}, CDXML\cite{cdxml}, and Molfile\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits. Like these systems, Dialect augments its underlying graph with node and edge labels.

@@

Two attributes characterize an atom's nucleus: \ttt{element} and \ttt{isotope}. The `element` attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC.\cite{periodicTable} When IUPAC authorizes additional symbols, they will become valid values for the \ttt{element} attribute. If the elemental identity of an atom is unknown, its \ttt{element} attribute can be omitted. The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, where mass number is the sum of proton and neutron count. Omitting the \ttt{isotope} property means that the isotopic composition is unspecified. When present, the lower bound on \ttt{isotope} equals one. This constraint leaves room for physically meaningless negative implied neutron count as in, for example, carbon-5. The lower bound for isotope applies even if the \ttt{element} attribute is not defined.

@@

The remaining constitutional atomic attribute, \ttt{charge}, enables electron counting. Electron counting associates each electron in a molecular graph with either an atom or a bond. This is a fundamental requirement for the principle of mass conversion. In a more practical sense, electron counting lays the foundation for calculations of formal charge and bond order.\cite{dietz:1995}

Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model).\cite{lewis:1916} Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly. The perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.

@@

A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME). DIME occurs when a molecular graph exists in two or more unequal forms, each differing from the others only through electron delocalization. Two molecular graphs are equal if they are isomorphic and a 1:1 mapping exists between the atoms and bonds in each graph. Electron delocalization results when the electrons of a valence bond representation can be distributed over more than one pair of atoms, as is the case with benzene and related molecules. DIME is similar to "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity.\cite{randic:2018,kerber:2006}

@@

A filled DS can be emptied through a two-step process of \textit{deselection}. First, a perfect matching over the DS is found. Next, each matched edge is replaced by a double bond. Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds. A popular algorithm for matching, the Edmunds "blossom algorithm,"\cite{edmonds:1965} has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes. Although more efficient algorithms are known, they are either much more difficult to implement or lack generality.

@@

Not all tetracoordinate atoms will be \textit{stereocenters}. A stereocenter is an atom whose "ligand permutation produces stereoisomers," as defined by Mislow and Siegel.\cite{mislow:1984} In the context of Dialect, a stereocenter must be tetracoordinate. Readers should avoid the assignment of stereodescriptors to atoms that are not stereocenters.

Special handling is required for \textit{undefined stereocenters}. A stereocenter is undefined if it lacks a stereodescriptor. Omitting a stereodescriptor conveys to readers that no information about the stereocenter's configuration is known. This could imply that either a single configuration or a configurational mixture are present. This interpretation is consistent with the one used by molfile format.\cite{ctfileFormats}

@@

Strings conform to an \textit{LL(1) grammar}. An LL(1) grammar is a context-free grammar whose strings can be parsed one character at a time from left to right with at most one character of lookahead. Additionally, LL(1) grammars expand the leftmost non-terminal first. This set of features makes LL(1) grammars such as the one used by Dialect a good fit for manually-written recursive descent parsers. LL(1) grammars can also be used as a basis for auto-generated parsers through packages such as ANTLR.\cite{parr:2014} The full grammar for Dialect strings is available as a text file in this paper's Supporting Material.

@@

Language subsets have ample precedent in computer science. A well-known example is MISRA C,\cite{misraC} a subset of the C programming language that aims to eliminate "known undefined or otherwise dangerous behavior."\cite{hatton:2007} Another example is the subset of JavaScript described by Crockford whose purpose is to "chip away at the features that are not beautiful until the language's true nature reveals itself."\cite{crockford:2008} Language subsets arise for two main reasons. First, designing a language presents many opportunities for errors small and large. Second, popular languages inevitably bring with them many users with diverse problems that demand new features. Some of these features cause future problems of their own. A language subset can improve both situations by eliminating the most problematic features.

@@

\begin{itemize}
    \item Weininger's report to the US Environmental Protection Agency ("The Article").\cite{anderson:1987}
    \item Weininger's original publication (herein "The Article").\cite{weininger:1988}
    \item Weininger's 2003 book chapter, revised in 2008 ("The Book Chapter").\cite{weininger:2008}
    \item An online manual maintained by Daylight ("The Manual").
    \item The Daylight's toolkit, which implements a SMILES reader and writer ("The Toolkit").\cite{daylightToolkit}
\end{itemize}

@@

The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete.\cite[p.~83]{weininger:2008} The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior.\cite{sageson:1997} Accordingly, Dialect disallows the symbol Ha. To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og. Moreover, Dialect explicitly accepts all future element symbols approved by IUPAC. The addition of these element symbols means that Dialect can't be considered a strict subset of SMILES. The alternative would, however, leave dialect without the ability to express every kind of IUPAC-recognized element.

The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules...".\cite[p.~33]{weininger:1988} The comma character is explicitly disallowed in Dialect.

The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses.\cite[p.~86]{weininger:2008} Dialect disallows such constructs, including for example "*((*))*" and "*(((*)))*".

"Reaction SMILES" is an extension described in the Book Chapter,\cite[p.~89]{weininger:2008} but disallowed in Dialect. Specifically, the greater than symbol (\ttt{>}) is not a valid Dialect character. This precludes strings such as "*>>*" in Dialect. Furthermore, Dialect lacks the "map" attribute, used for atom-atom mapping in reactions, and its accompanying syntax.

Dialect only supports two stereodescriptors, encoded with the terminals "@" and "@@". The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@".\cite[p.~94]{weininger:2008} The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1". None of these are supported by Dialect. Nor does Dialect support the application of tetracoordinate stereodescriptors to odd cumulenes as described in the book chapter. Although some of these forms of configuration are supported today, it's unusual to find them all supported. The reason is simple: in the context in which Dialect will most often used, these forms of stereochemistry are rare. To support them burdens implementors with little payoff.

ProtoSMILES supports the concept of "aromatic bond" (":"), but Dialect does not. Despite clear inclusion within the syntax of ProtoSMILES, no precise guidance on using this feature is provided. The Book Chapter notes "Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond). "-" (single) and ":" (aromatic) bond symbols may always be omitted on input".\cite[p.~85]{weininger:2008} This passage implies that aromatic bonds serve no purpose. Accordingly, Dialect does not allow them and the syntax does not include the colon symbol (\ttt{:}).

Along these lines, Dialect does not recognize atoms bearing either arsenic (As) or selenium (Se) as selectable. Doing so would increase the complexity of the language unnecessarily. Although The Paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context."),\cite[p.~35]{weininger:1988} default valences for the latter two atoms are not provided. Even if they were, the elements As and Se must be bracketed. As such, special rules for handling them in the context of pruning would be required. Moreover, molecules bearing the elements As or Se are often made "aromatic" unnecessarily in SMILES. Finally, allowing the selection of atoms bearing As or Se does nothing to address molecules with atoms bearing silicon, germanium, tin, or other heavy main group elements that may be discovered in the future to participate in DIME.

It's unclear whether ProtoSMILES allows a virtual hydrogen count on hydrogen itself. The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence ("There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].").\cite[p.~34]{weininger:1988} If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as "[HH]". The Book Chapter appears to reiterate this view, noting that "explicit hydrogen specification is required" in the case of "hydrogens connected to other hydrogens, e.g. [H][H], molecular hydrogen".\cite[p.~97]{weininger:2008} To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself. For example, the following are all valid, if chemically nonsensical, strings: "[HH]"; "[HH2]"; and "[HH3]". Depending on interpretation, this feature may or may not render Dialect a superset of SMILES.

ProtoSMILES ascribes ambiguous meaning to detachments. As noted in the Book Chapter, "... In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other.".\cite[p.~88]{weininger:2008} The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent.\cite{clark:2011} To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.

ProtoSMILES assigns no explicit upper or lower bounds to numerical atomic attributes. These boundaries are nevertheless crucial for implementors, who seek the data types necessary to prevent underflow and overflow. For this reason, Dialect sets both upper and lower bounds on the following atomic attributes: isotope (0 $\leq$ value \textless 1000); charge (-10, \textless value \textless 10); and hydrogens, 0 $\leq$ value \textless 10). Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be implicit, the lack of precision forces implementors to improvise, which can lead to divergent behavior. ProtoSMILES places no upper bound on the charge notation "+", "++", "+++", and their corresponding negative forms ("The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3].").\cite[p.~32]{weininger:1988} However, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.

@@

Dialect also resolves some internal contradictions within ProtoSMILES. The Article reports the only default valence for nitrogen as three (p. 34),\cite[p.~34]{weininger:1988} whereas the Book Chapter reports two default valences: 3 and 5.\cite[p.~84]{weininger:2008} Dialect uses the latter default valences. Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I.\cite[p.~32]{weininger:2008} However, an example later in the Article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string.\cite[p.~35]{weininger:1988} Dialect explicitly rejects strings containing unbracketed hydrogen atoms. On the topic of atom selection ("aromaticity"), the Article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion."\cite[p.~34]{weininger:1988} The Book Chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc").\cite[p 85]{weininger:2008} Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.

>>>>

Simplified Molecular Input Line Entry System (SMILES) was first described by Weininger in 1988~\cite{weininger:1988}. As a line notation~\cite{wiswesser:1968}, SMILES represents molecules as single line character sequences, or strings. SMILES strings can be read and written algorithmically~\cite{weininger:1989,oBoyle:2012}, and can be canonicalized to yield unique molecular identifiers.

SMILES has since been widely adopted. Read and write functionalities are routinely supported by popular cheminformatics toolkits, including: Open Babel~\cite{oBoyle:2011}; RDKit~\cite{rdkit}; Chemistry Development Kit~\cite{steinbeck:2003}; JChem Base~\cite{jchemBase}; the Daylight Toolkit~\cite{daylightToolkit}; and OEChem TK~\cite{oetk}. SMILES encodings can be found in many public-facing databases, including: PubChem~\cite{kim:2016}; ChEBI~\cite{degtyarenko:2007}; ZINC~\cite{irwin:2005}; ChEMBL~\cite{gaulton:2012}; and Wikipedia~\cite{wikipediaCAS}. Raw SMILES strings have been used extensively in machine learning applications~\cite{sousa:2021}. SMILES has also been extended to carry various forms of metadata, as exemplified by Jmol SMILES~\cite{hanson:2016}, CurleySMILES~\cite{drefahl:2011}, BigSMILES~\cite{lin:2019}, and CXSMILES~\cite{cxsmiles}.

@@

Earlier descriptions of SMILES were published in various media, although these are not cited by Weininger. A currently-accessible example is a 1987 report to the US Environmental Protection Agency~\cite{anderson:1987}. Although the language described there shares many features with later descriptions, there are important differences as well.

Since 1988, two authoritative sources have attempted to clarify or revise Weininger's 1988 description of SMILES. A 2003 book chapter by Weininger addressed some of the limitations of the paper~\cite{weininger:2008}. The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc. ("Daylight"), recapitulates previously-published material~\cite{daylightTheory}. The lack of other written sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication~\cite{gasteiger:2018}.

In 2007 a documentation effort that would become known as OpenSMILES began~\cite{openSMILES}. OpenSMILES was conceived as "a non-proprietary specification for the SMILES language." Many of the points left open by previous SMILES documentation efforts were addressed. Noteworthy contributions included: the first formal grammar, which was later adapted~\cite{denjoed:2010}; many refinements around stereochemistry; and the introduction of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count. OpenSMILES also left several points of syntactic and semantic ambiguity unsettled, while introducing extensions of its own.

In 2019 IUPAC announced the SMILES+ initiative~\cite{smilesPlusSpecification}. Noting the limitations of existing SMILES documentation, the SMILES+ effort seeks to "establish a formalized recommended up-to-date specification of the SMILES format." SMILES+ took as its starting point the documentation produced by the OpenSMILES project. Efforts to extend this starting point are in progress online through a public repository, but no formal recommendation has to date resulted~\cite{smilesPlusGitHub}.

@@

The lack of a detailed, comprehensive, and universally-applied SMILES specification also hampers broader data integrity efforts. The FAIR Guiding Principles identify four qualities essential for extracting maximum utility from published scholarly data: Findability; Accessibility; Interoperability; and Reusability~\cite{wilkinson:2016}. Interoperability in particular requires that "(meta)data" use a formal, accessible, shared, and broadly applicable language for knowledge representation." Although SMILES may be a shared, accessible, and broadly applicable language, its formal underpinnings leave room for improvement.

This situation has persisted long enough that the notion of a SMILES "standard" may not currently be feasible. A dozen or more slightly different implementations are now in use. None of them can be considered a reference standard. Some have introduced extensions not recognized by the others, contributing to language "drift" while compounding the already formidable difficulties faced by implementors. Others claim SMILES compatibility while ignoring certain aspects of written documentation. The Daylight implementation, like some of the others, is distributed under a non-Open Source license that restricts use~\cite{daylightToolkit}. For several years Daylight operated a publicly-accessible web service based on its toolkit that depicted SMILES strings, but it has since been decommissioned~\cite{depictArchive}. Although SMILES may underpin modern chemical information exchange, the foundations of this system are weathered and showing signs of stress.

@@

Dialect's system of representation is based on the \textit{molecular graph} concept~\cite{balaban:1985}. A graph is a data structure comprised of a set of nodes (which map to "atoms") and a set of pairwise relationships between them called edges (which map to "bonds"). A molecular graph is a specialized graph onto which atomic and bonding metadata have been overlaid. Molecular graphs are ubiquitous in cheminformatics, appearing in contexts such as data formats like Chemical Markup Language~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and Molfile~\cite{ctfileFormats}, as well as in-memory data structures found in cheminformatics toolkits. Like these systems, Dialect augments its underlying graph with node and edge labels.

@@

Two attributes characterize an atom's nucleus: \ttt{element} and \ttt{isotope}. The `element` attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC~\cite{periodicTable}. When IUPAC authorizes additional symbols, they will become valid values for the \ttt{element} attribute. If the elemental identity of an atom is unknown, its \ttt{element} attribute can be omitted. The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, where mass number is the sum of proton and neutron count. Omitting the \ttt{isotope} property means that the isotopic composition is unspecified. When present, the lower bound on \ttt{isotope} equals one. This constraint leaves room for physically meaningless negative implied neutron count as in, for example, carbon-5. The lower bound for isotope applies even if the \ttt{element} attribute is not defined.

@@

The remaining constitutional atomic attribute, \ttt{charge}, enables electron counting. Electron counting associates each electron in a molecular graph with either an atom or a bond. This is a fundamental requirement for the principle of mass conversion. In a more practical sense, electron counting lays the foundation for calculations of formal charge and bond order~\cite{dietz:1995}.

Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model)~\cite{lewis:1916}. Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly. The perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.

@@

A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME). DIME occurs when a molecular graph exists in two or more unequal forms, each differing from the others only through electron delocalization. Two molecular graphs are equal if they are isomorphic and a 1:1 mapping exists between the atoms and bonds in each graph. Electron delocalization results when the electrons of a valence bond representation can be distributed over more than one pair of atoms, as is the case with benzene and related molecules. DIME is similar to "resonance" or "aromaticity," but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.

@@

A filled DS can be emptied through a two-step process of \textit{deselection}. First, a perfect matching over the DS is found. Next, each matched edge is replaced by a double bond. Because the presence of a filled DS implies a perfect matching over it, kekulization always succeeds. A popular algorithm for matching, the Edmunds "blossom algorithm," has a time complexity of $O(|E||V|^2)$, where $|E|$ is the number of edges and $|V|$ is the number of nodes~\cite{edmonds:1965}. Although more efficient algorithms are known, they are either much more difficult to implement or lack generality.

@@

Not all tetracoordinate atoms will be \textit{stereocenters}. A stereocenter is an atom whose "ligand permutation produces stereoisomers," as defined by Mislow and Siegel~\cite{mislow:1984}. In the context of Dialect, a stereocenter must be tetracoordinate. Readers should avoid the assignment of stereodescriptors to atoms that are not stereocenters.

Special handling is required for \textit{undefined stereocenters}. A stereocenter is undefined if it lacks a stereodescriptor. Omitting a stereodescriptor conveys to readers that no information about the stereocenter's configuration is known. This could imply that either a single configuration or a configurational mixture are present. This interpretation is consistent with the one used by molfile format~\cite{ctfileFormats}.

@@

Strings conform to an \textit{LL(1) grammar}. An LL(1) grammar is a context-free grammar whose strings can be parsed one character at a time from left to right with at most one character of lookahead. Additionally, LL(1) grammars expand the leftmost non-terminal first. This set of features makes LL(1) grammars such as the one used by Dialect a good fit for manually-written recursive descent parsers. LL(1) grammars can also be used as a basis for auto-generated parsers through packages such as ANTLR~\cite{parr:2014}. The full grammar for Dialect strings is available as a text file in this paper's Supporting Material.

@@

Language subsets have ample precedent in computer science. A well-known example is MISRA C~\cite{misraC}, a subset of the C programming language that aims to eliminate "known undefined or otherwise dangerous behavior"~\cite{hatton:2007}. Another example is the subset of JavaScript described by Crockford whose purpose is to "chip away at the features that are not beautiful until the language's true nature reveals itself."~\cite{crockford:2008}. Language subsets arise for two main reasons. First, designing a language presents many opportunities for errors small and large. Second, popular languages inevitably bring with them many users with diverse problems that demand new features. Some of these features cause future problems of their own. A language subset can improve both situations by eliminating the most problematic features.

@@

\begin{itemize}
    \item Weininger's report to the US Environmental Protection Agency ("The Article")~\cite{anderson:1987}.
    \item Weininger's original publication (herein "The Article")~\cite{weininger:1988}.
    \item Weininger's 2003 book chapter, revised in 2008 ("The Book Chapter")~\cite{weininger:2008}.
    \item An online manual maintained by Daylight ("The Manual").
    \item The Daylight's toolkit, which implements a SMILES reader and writer ("The Toolkit")~\cite{daylightToolkit}.
\end{itemize}

@@

The Book Chapter defines a fixed set of element symbols which even at the time of publication was obsolete~\cite[p.~83]{weininger:2008}. The symbol for element 105 is given as Ha (Hahnium), despite IUPAC's resolution of this naming controversy in favor of the symbol Db (Dubnium) years prior~\cite{sageson:1997}. Accordingly, Dialect disallows the symbol Ha. To support the full set of currently-named transuranium elements, Dialect adds the symbols: Db; Sg; Bh; Hs; Mt; Rg; Cn; Nh; Fl; Mc; Lv; Ts; and Og. Moreover, Dialect explicitly accepts all future element symbols approved by IUPAC. The addition of these element symbols means that Dialect can't be considered a strict subset of SMILES. The alternative would, however, leave dialect without the ability to express every kind of IUPAC-recognized element.

The Article implies that comma (\ttt{,}) is either an element symbol or a valid SMILES character: "This subset uses only the symbols H, C, N, O, P, S, F, Cl, Br, I, and (,) and digits, with the following four rules..."~\cite[p.~33]{weininger:1988}. The comma character is explicitly disallowed in Dialect.

The Book Chapter defines a recursive grammar for branching which, perhaps without the knowledge of the author, defines sequential parentheses~\cite[p.~86]{weininger:2008}. Dialect disallows such constructs, including for example "*((*))*" and "*(((*)))*".

"Reaction SMILES" is an extension described in the Book Chapter~\cite[p.~89]{weininger:2008}, but disallowed in Dialect. Specifically, the greater than symbol (\ttt{>}) is not a valid Dialect character. This precludes strings such as "*>>*" in Dialect. Furthermore, Dialect lacks the "map" attribute, used for atom-atom mapping in reactions, and its accompanying syntax.

Dialect only supports two stereodescriptors, encoded with the terminals "@" and "@@". The Book Chapter provides a recursive grammar for stereodescriptors that allows multi-symbol descriptors such as "@@@" and "@@@@"~\cite[p.~94]{weininger:2008}. The book chapter also discusses non-tetrahedral descriptors including "@AL1", "@AL2", "@1", and "@SP1". None of these are supported by Dialect. Nor does Dialect support the application of tetracoordinate stereodescriptors to odd cumulenes as described in the book chapter. Although some of these forms of configuration are supported today, it's unusual to find them all supported. The reason is simple: in the context in which Dialect will most often used, these forms of stereochemistry are rare. To support them burdens implementors with little payoff.

ProtoSMILES supports the concept of "aromatic bond" (":"), but Dialect does not. Despite clear inclusion within the syntax of ProtoSMILES, no precise guidance on using this feature is provided. The Book Chapter notes "Adjacent atoms without an intervening bond symbol are connected by a valence-dictated bond (typically a single or aromatic bond). "-" (single) and ":" (aromatic) bond symbols may always be omitted on input"~\cite[p.~85]{weininger:2008}. This passage implies that aromatic bonds serve no purpose. Accordingly, Dialect does not allow them and the syntax does not include the colon symbol (\ttt{:}).

Along these lines, Dialect does not recognize atoms bearing either arsenic (As) or selenium (Se) as selectable. Doing so would increase the complexity of the language unnecessarily. Although The Paper implies selectability ("Elements other than C, N, O, P, S, As, and Se are not yet dealt with in an aromatic context.")~\cite[p.~35]{weininger:1988}, default valences for the latter two atoms are not provided. Even if they were, the elements As and Se must be bracketed. As such, special rules for handling them in the context of pruning would be required. Moreover, molecules bearing the elements As or Se are often made "aromatic" unnecessarily in SMILES. Finally, allowing the selection of atoms bearing As or Se does nothing to address molecules with atoms bearing silicon, germanium, tin, or other heavy main group elements that may be discovered in the future to participate in DIME.

It's unclear whether ProtoSMILES allows a virtual hydrogen count on hydrogen itself. The Article implies that the element symbol H (hydrogen) must not have an implicit hydrogen count when used within a bracket sequence ("There are few exceptions to the hydrogen-suppression convention, the most obvious being specification of a proton, [H+], and molecular hydrogen, [H][H].")~\cite[p.~34]{weininger:1988}. If so, the author may have erroneously believed that parsers would be incapable of processing constructs such as "[HH]". The Book Chapter appears to reiterate this view, noting that "explicit hydrogen specification is required" in the case of "hydrogens connected to other hydrogens, e.g. [H][H], molecular hydrogen"~\cite[p.~97]{weininger:2008}. To be clear, Dialect supports any valid virtual hydrogen count on hydrogen itself. For example, the following are all valid, if chemically nonsensical, strings: "[HH]"; "[HH2]"; and "[HH3]". Depending on interpretation, this feature may or may not render Dialect a superset of SMILES.

ProtoSMILES ascribes ambiguous meaning to detachments. As noted in the Book Chapter, "... In terms of the valence model being represented, the dot literally represents a bond of formal order zero: the atoms on either side of the dot are explicitly not bonded to each other."~\cite[p.~88]{weininger:2008}. The state of not being bonded and the state of having a zero-order bond are not necessarily equivalent~\cite{clark:2011}. To be clear, Dialect explicitly disallows the zero bond order interpretation of detachments and the dot character.

ProtoSMILES assigns no explicit upper or lower bounds to numerical atomic attributes. These boundaries are nevertheless crucial for implementors, who seek the data types necessary to prevent underflow and overflow. For this reason, Dialect sets both upper and lower bounds on the following atomic attributes: isotope (0 $\leq$ value \textless 1000); charge (-10, \textless value \textless 10); and hydrogens, 0 $\leq$ value \textless 10). Although it might be argued that lower bounds on physical quantities such as isotope and hydrogen count should be implicit, the lack of precision forces implementors to improvise, which can lead to divergent behavior. ProtoSMILES places no upper bound on the charge notation "+", "++", "+++", and their corresponding negative forms ("The SMILES program also recognizes constructions of the form [Fe+++] as being synonymous with the form [Fe+3].")~\cite[p.~32]{weininger:1988}. However, Dialect disallows duplicated plus and minus characters (\ttt{+} and \ttt{-}, respectively) from bracket atoms.

@@

Dialect also resolves some internal contradictions within ProtoSMILES. The Article reports the only default valence for nitrogen as three (p. 34)~\cite[p.~34]{weininger:1988}, whereas the Book Chapter reports two default valences: 3 and 5~\cite[p.~84]{weininger:2008}. Dialect uses the latter default valences. Atoms whose \ttt{hydrogens} attribute may be set to \ttt{Implicit} ("the organic subset") are reported in the Article as: B; C; N; O; P; S; F; Cl; Br; and I~\cite[p.~32]{weininger:2008}. However, an example later in the Article indicates that "Hn1cccc1" with its unbracketed hydrogen is a valid string~\cite[p.~35]{weininger:1988}. Dialect explicitly rejects strings containing unbracketed hydrogen atoms. On the topic of atom selection ("aromaticity"), the Article implies that selected atoms must be part of a cycle: "To qualify as aromatic, all atoms in the ring must be sp2 hybridized and the number of available 'excess' $\pi$ electrons must satisfy Hückel's 4{\textit{N}} + 2 criterion."~\cite[p.~34]{weininger:1988}. The Book Chapter, however, presents two examples of valid strings that do not encode cycles ("cc" and "cccc")~\cite[p 85]{weininger:2008}. Consistent with the latter view, Dialect explicitly allows the selection of atoms regardless of cycle membership.
